
\begin{thebibliography}{AAAA00}

\bibitem[Atk91]{Atk91}
  A.O.L.~Atkin.
  \newblock \emph{The number of points on an elliptic curve modulo a prime}.
  \newblock Email on the Number Theory Mailing List, 1991.

\bibitem[BK78]{BrKu78}R.P.~Brent and H.T.~Kung, Fast algorithms for
  manipulating formal power series.  \emph{J. ACM} 25(4):581--595, 1978.

\bibitem[BSS99]{BlSeSm99}
  I.~Blake, G.~Seroussi, and N.~Smart.
  \newblock {\em Elliptic curves in cryptography}.
  \newblock Cambridge University Press, 1999.

\bibitem[BMSS08]{BoMoSaSc08}
  A.~Bostan, F.~Morain, B.~Salvy and É.~Schost
  \newblock Fast algorithms for computing isogenies between elliptic curves.
  \newblock \emph{Math. Comp.} 77, 263:1755-1778, 2008.

\bibitem[Magma]{Magma}
  W.~Bosma, J.~Cannon, C.~Playoust.
  \newblock The Magma algebra system. I. The user language.
  \newblock \emph{J. Symb. Comp.}, 24(3-4):235-265, 1997.
  
\bibitem[gf2x]{gf2x}
  R.~Brent, P.~Gaudry, E.~Thom{\'e}, P.~Zimmermann.
  \newblock Faster multi\-plication in GF$(2)[x]$.
  \newblock In {\em ANTS'08}, 153--166. Springer, 2008.

\bibitem[Cou94]{Cou94}
  J.-M.~Couveignes.
  \newblock \emph{Quelques calculs en théorie des nombres}.
  \newblock PhD thesis. 1994.

\bibitem[Cou96]{Cou96}
  J.-M.~Couveignes. 
  \newblock Computing {$\ell$}-isogenies using the {$p$}-torsion.
  \newblock in {\em ANTS'II}, 59--65. Springer, 1996.

\bibitem[Cou00]{Cou00}
  J.-M.~Couveignes. 
  \newblock Isomorphisms between {A}rtin-{S}chreier towers. 
  \newblock \emph{Math. Comp.} 69(232): 1625--1631, 2000.
  
\bibitem[DFS10]{DFS09}
  L.~De~Feo and \'E.~Schost
  \newblock Fast Arithmetics in Artin Schreier Towers.
  \newblock \emph{Preprint}, 2010.

\bibitem[Elk98]{Elk91}
  N.D.~Elikes
  \newblock Elliptic and modular curves over finite fields and related computational issues.
  \newblock \emph{Computational Perspectives on Number Theory: Proceedings of a Conference in Honor of A.O.L. Atkin}, 21--76, AMS, 1998. 

\bibitem[EM03]{EnMo03}
  A.~Enge and F.~Morain.
  \newblock Fast decomposition of polynomials with known Galois group.
  \newblock in {\em AAECC-15}, 254--264. Springer, 2003.

\bibitem[JL06]{JL06}A.~Joux, R. Lercier.
  \newblock Counting points on elliptic curves in medium characteristic.
  \newblock \emph{Cryptology ePrint Archive} 2006/176, 2006.

\bibitem[vzGG]{vzGG} 
  J.~von~zur~Gathen and J.~Gerhard. 
  \newblock \emph{Modern Computer Algebra}. 
  \newblock Cambridge University Press, 1999.

\bibitem[vzGS92]{vzGS92} 
  J.~von~zur~Gathen and V.~Shoup. 
  \newblock Computing Frobenius maps and factoring polynomials
  \newblock \emph{Comput. Complexity}, vol. 2, 187--224, 1992.
  
\bibitem[Gun76]{Gun76}H.~Gunji.
    \newblock The Hasse Invariant and $p$-division Points of an Elliptic Curve.
  \newblock \emph{Archiv der Mathematik} 27(2), Springer, 1976.

\bibitem[KS97]{KS97}
  E.~Kaltofen and V.~Shoup.
  \newblock Fast polynomial factorization over high algebraic extensions
  of finite fields.
  \newblock In \emph{ISSAC '97}, 184--188. ACM, 1997.

\bibitem[KS98]{KS98}
  E.~Kaltofen and V.~Shoup.
  \newblock Subquadratic-time factoring of polynomials over finite fields.
  \newblock \emph{Math. Comput.}, 1179--1197, AMS, 1998.

\bibitem[KU08]{KeUm08}
  K.~S. Kedlaya and C.~Umans
  \newblock Fast modular composition in any characteristic
  \newblock In {\em FOCS'08}, 146--155, IEEE, 2008
  
\bibitem[Ler96]{Ler96}
  R.~Lercier.
  \newblock Computing isogenies in GF($2^n$).
  \newblock In {\em ANTS-II}, LNCS vol 1122, 197--212. Springer, 1996.

\bibitem[Ler97]{Ler97}
  R.~Lercier.
  \newblock Algorithmique des courbes elliptiques dans les corps finis. 
  \newblock Ph.D. Thesis, {\'E}cole polytechnique, 1997.

\bibitem[LS09]{LeSi09}
  R.~Lercier, T.~Sirvent.
  \newblock On Elkies subgroups of $\ell$-torsion points in curves defined over
  a finite field.
  \newblock To appear in {\em J. Th\'eor. Nombres Bordeaux}.

\bibitem[Mon87]{Mon87}
  P.~L.~Montgomery
  \newblock Speeding the Pollard and Elliptic Curve Methods of Factorization
  \newblock \emph{Math. Comp.}, Vol. 48, No. 177., 243--264, 1987.

\bibitem[RS06]{RoSt06}
  A.~Rostovtsev and A.~Stolbunov.
  \newblock Public-key cryptosystem based on isogenies.
  \newblock Cryptology ePrint Archive, Report 2006/145.
  
\bibitem[Sat00]{Sat00}
  T.~Satoh.
  \newblock The canonical lift of an ordinary elliptic curve over a nite eld and its point counting.
  \newblock \emph{Journal of the Ramanujan Mathematical Society}, 2000.

\bibitem[Sch95]{Sch95}
  R.~Schoof.
  \newblock Counting points on elliptic curves over finite fields.
  \newblock \emph{J. Théorie des Nombres de Bordeaux} 7:219--254, 1995.

\bibitem[NTL]{NTL}
  V.~Shoup.
  \newblock {NTL}: A library for doing number theory.
  \newblock \url{http://www.shoup.net/ntl/}.

\bibitem[Tes06]{Tes06}
  E.~Teske.
  \newblock An elliptic trapdoor system.
  \newblock {\em Journal of Cryptology}, 19(1):115--133, 2006.

\bibitem[V\'el71]{Vel71}J.~V\'elu.
  \newblock Isogénies entre courbes elliptiques.
  \newblock \emph{Comptes Rendus de l'Académie des Sciences de Paris} 273,
  Série A, 238--241, 1971.

\bibitem[Vol90]{Vol90}J.F.~Voloch.
  \newblock Explicit $p$-descent for Elliptic Curves in Characteristic $p$.
  \newblock \emph{Compositio Mathematica} 74, 247--58, 1990.
  
\end{thebibliography}
