\section{The algorithm C2}
\label{sec:C2}

The algorithm we refer to as C2 was originally proposed in
\cite{Cou96}. It takes as input two elliptic curves $E, E'$ and an
integer $\ell$ prime to $p$ and it returns, if it exists, an
$\F_q$-rational isogeny of degree $\ell$ between $E$ and $E'$. It only
works in odd characteristic.

\subsection{The original algorithm}
Suppose there exists an $\F_q$-rational isogeny
$\I:E\rightarrow E'$ of degree $\ell$. Since $\ell$ is prime to $p$
one has $\I(E[p^k]) = E'[p^k]$ for any $k$.

Recall that $E[p^k]$ and $E'[p^k]$ are cyclic groups. C2 iteratively
computes generators $P_k,P_k'$ of $E[p^k]$ and $E'[p^k]$
respectively. Now C2 makes the guess $\I(P_k) = P_k'$; then, if $\I$
is given by rational fractions as in \eqref{eq:isog},
\begin{equation}
  \label{eq:C2:I}
  \frac{g\bigl(x([i]P_k)\bigr)}{h\bigl(x([i]P_k)\bigr)} = x([i]P_k')
  \quad\text{for $i\in\Z/p^k\Z$} 
\end{equation}
and by \eqref{eq:velu-deg} $\deg g = \deg h + 1 = \ell$.

Using \eqref{eq:C2:I} one can compute the rational fraction
$\frac{g(X)}{h(X)}$ through Cauchy interpolation over the points of
$E[p^k]$ for $k$ large enough. C2 takes $p^k > 4\ell - 2$,
interpolates the rational fraction and then checks that it corresponds
to the restriction of an isogeny to the $x$-axis. If this is the case,
the whole isogeny is computed through Vélu formulae and the algorithm
terminates. Otherwise the guess $\I(P_k) = P_k'$ was wrong, then C2
computes a new generator for $E'[p^k]$ and starts over again.

We now go through the details of the algorithm.

\paragraph{The $p$-torsion}
The computation of the $p$-torsion points follows from the work of
Gunji \cite{Gun76}. Here we suppose $p\ne2$.

\begin{definition}
  \label{def:hasse}
  Let $E$ have equation $y^2 = f(x)$. The \emph{Hasse invariant} of
  $E$, noted $H_E$, is the coefficient of $X^{p-1}$ in
  $f(X)^{\frac{p-1}{2}}$.
\end{definition}

Gunji shows the following proposition and gives formulae to compute
the $p$-torsion points.

\begin{proposition}
  \label{th:gunji}
  Let $c=\sqrt[p-1]{H_E}$; then, the $p$-torsion points of $E$ are
  defined in $\F_q[c]$ and their abscissae are defined in $\F_q[c^2]$.
\end{proposition}


\paragraph{The $p^k$-torsion}
$p^k$-torsion points are iteratively computed via $p$-descent. The
basic idea is to split the multiplication map as $[p] = \frobisog\circ
V$ and invert each of the components. The purely inseparable isogeny
$\frobisog$ is just a frobenius map and the separable isogeny $V$ can
be computed by Vélu formulae once the $p$-torsion points are
known. Although this is reasonably efficient, pulling $V$ back may
involve factoring polynomials of degree $p$ in some extension field.

A finer way to do the $p$-descent, as suggested in the original paper
\cite{Cou96}, is to use the work of Voloch \cite{Vol90}. Suppose
$p\ne2$, let $E$ and $\widetilde{E}$ have equations respectively
\begin{align*}
  y^2&=f(x)=x^3+a_2x^2+a_4x+a_6 \;\text{,}\\
  \tilde{y}^2&=\tilde{f}(\tilde{x}) = \tilde{x}^3 +
  \sqrt[p]{a_2}\tilde{x}^2 + \sqrt[p]{a_4}\tilde{x} + \sqrt[p]{a_6}
  \;\text{,}
\end{align*}
set
 \begin{equation}
  \label{eq:voloch:cover}
  \tilde{f}(X)^{\frac{p-1}{2}} = \alpha(X) + H_{\widetilde{E}}X^{p-1} + X^p\beta(X)
\end{equation}
with $\deg \alpha < p-1$ and $H_{\widetilde{E}}$ the Hasse invariant
of $\widetilde{E}$. Voloch shows the following proposition.

\begin{proposition}
  \label{th:voloch}
  Let $\tilde{c} = \sqrt[p-1]{H_{\widetilde{E}}}$, the cover of
  $\widetilde{E}$ defined by
  \begin{equation}
    \label{th:voloch:cover}
    C:\; \tilde{z}^p - \tilde{z} = \frac{\tilde{y}\beta(\tilde{x})}{\tilde{c}^p}
  \end{equation}
  is an étale cover of degree $p$ and is isomorphic to $E$ over
  $\F_q[\tilde{c}]$; the isomorphism is given by
  \begin{equation}
    \label{th:voloch:isom}
    \left\{
      \begin{aligned}
        (\tilde{x}, \tilde{y}) &= V(x, y)\\
        \tilde{z} &= -\frac{y}{\tilde{c}^p}\sum_{i=1}^{p-1}\frac{1}{x - x([i]P_1)}
      \end{aligned}
    \right.
  \end{equation}
  where $P_1$ is a primitive $p$-torsion point of $E$.
\end{proposition}

The descent is then performed as follows: starting from a point $P$ on
$E$, first pull it back along $\frobisog$, then take one of its
pre-images in $C$ by solving equation \eqref{th:voloch:cover}, finally
use equation \eqref{th:voloch:isom} to land on a point $P'$ in $E$.
The proposition guarantees that $[p]P' = P$. The descent is pictured
in figure \ref{fig:voloch}.

\begin{figure}
  \centering
  \[
  \xymatrix{\widetilde{E}\ar@/^/[r]^{\frobisog} & E\ar@/^/[l]^{V}}
  %%
  \qquad
  %%
  \xymatrix{
    \widetilde{E}\ar@/^/[r]^{\frobisog} & E\ar@/^/@{-->}[l]^{V}\ar[d]_{\simeq}\\
    & C\ar@/^/[ul]
  }
  \]
  
  \caption{Two ways of doing the $p$-descent: standard on the left and via a degree $p$ cover on the right}
  \label{fig:voloch}
\end{figure}


The reason why this is more efficient than a standard descent is the
shape of equation \eqref{th:voloch:cover}: it is an Artin-Schreier
equation and it can be solved by many techniques, the simplest being
linear algebra (as was suggested in \cite{Cou96}). Once a solution
$\tilde{z}$ to \eqref{th:voloch:cover} is known, solving in $x$ and $y$ the
bivariate polynomial system \eqref{th:voloch:isom} takes just a GCD
computation (explicit formulae were given by Lercier in
\cite[$\S$6.2]{Ler97}, we give some slightly improved ones in Section
\ref{sec:implementation}). Compare this with a generic factoring
algorithm needed by standard descent.

Solving Artin-Schreier equations is the most delicate task of the
descent and we will further discuss it.


\paragraph{Cauchy interpolation}
Interpolation reconstructs a polynomial from the values it takes on
some points; Cauchy interpolation reconstructs a rational
fraction. The Cauchy interpolation algorithm is divided in two phases:
first find the polynomial $P$ interpolating the evaluation points,
then use rational fraction reconstruction to find a rational fraction
congruent to $P$ modulo the polynomial vanishing on the points. The
first phase is carried out through any classical interpolation
algorithm, while the second is similar to an XGCD computation. See
\cite[$\S$5.8]{vzGG} for details.

Cauchy interpolation needs $n+2$ points to reconstruct a degree
$(k,n-k)$ rational fraction. This, together with \eqref{eq:velu-deg},
justifies the choice of $k$ such that $p^k > 4\ell - 2$. Some of our
variants of C2 will interpolate only on the primitive $p^k$-torsion
points, thus requiring the slightly larger bound $\euler(p^k) \ge
4\ell - 2$. This is not very important to our asymptotical analysis
since in both cases $p^k \in O(\ell)$.

\paragraph{Recognising the isogeny}
Once the rational fraction $\frac{g(X)}{h(X)}$ has been computed, one
has to verify that it is indeed an isogeny. The first test is to check
that the degrees of $g$ and $h$ match equation \eqref{eq:velu-deg}, if
they don't, the equation can be discarded right away and the algorithm
can go on with the next trial. Next, one can check that $h$ is indeed
the square of a polynomial (or, if $\ell$ is even, the product of one
factor of the $2$-division polynomial and a square polynomial). This
two tests are usually enough to detect an isogeny, but, should they
lie, one can still check that the resulting rational function is
indeed a group morphism by trying some random points on $E$.


\subsection{The case $p=2$}
\label{sec:p=2}
The algorithm as we have presented it only works when $p\ne2$, it is
however an easy matter to generalise it. The only phase that doesn't
work is the computation of the $p^k$-torsion points. For curves in the
form \eqref{eq:weierstrass=2} the only $2$-torsion point is
$(0,\sqrt{b})$.

Voloch formulae are hard to adapt, nevertheless a $2$-descent on the
Kummer surface of $E$ can easily be performed since the doubling
formula reads
\begin{equation}
  x([2]P) = \frac{b}{x(P)^2} + x(P)^2 =
  \frobisog\left(\frac{\sqrt{b} + x(P)^2}{x(P)} \right) = \frobisog\circ V
  \;\text{.}
\end{equation}
Given point $x_P$ on $K_E$, a pull-back along $\frobisog$ gives a
point $\tilde{x}_P$ on $K_{\widetilde{E}}$. Then pulling $V$ back
amounts to solve
\begin{equation}
  \label{eq:2-descent}
  x^2 + \tilde{x}_Px = \sqrt{b}
\end{equation}
and this can be turned in an Artin-Schreier equation through the
change of variables $x \rightarrow x'\tilde{x}_P$.

From the descent on the Kummer surfaces one could deduce a full
$2$-descent on the curves by solving a quadratic equation at each step
in order recover the $y$ coordinate, but this would be too
expensive. Fortunately, the $y$ coordinates are not needed by the
subsequent steps of the algorithm, thus one may simply ignore
them. Observe in fact that even if $K_E$ does not have a group law,
the restriction of scalar multiplication is well defined and can be
computed through Montgomery formulae \cite{Mon87}. This is enough to
compute all the abscissae of the points in $E[p^k]$ once a generator
is known.


\subsection{Complexity analysis}
\label{sec:C2:complexity}
Analysing the complexity of C2 is a delicate matter since the
algorithm relies on some black-box computer algebra algorithms in
order to deal with finite extensions of $\F_q$. The choice of the
actual algorithms may strongly influence the overall complexity of C2.
In this section we will only give some lower bounds on the complexity
of C2, since a much more accurate complexity analysis will be carried
out in Section \ref{sec:C2-AS}.

\paragraph{$p$-torsion}
Applying Gunji formulae first requires to find $c$ and $c'$, $p-1$-th
roots of $H_E$ and $H_{E'}$, and build the field extension $\F_q[c] =
\F_q[c']$. Independently of the actual algorithm used, observe that in
the worst case $\F_q[c]$ is a degree $p-1$ extension of $\F_q$, thus
simply representing one of its elements requires $\Theta(pd)$ elements
of $\F_p$.

Subsequently, the main cost in Gunji's formulae is the computation of
the determinant of a $\frac{p-1}{2}\times\frac{p-1}{2}$
quadri-diagonal matrix (see \cite{Gun76}). This takes $\Theta(p^2)$
operations in $\F_q[c]$ by Gauss elimination, that is no less than
$\Omega(p^3d)$ operations in $\F_p$.

\paragraph{$p^k$-torsion}
During the $p$-descent, factoring of equations \eqref{th:voloch:cover}
or \eqref{eq:2-descent} may introduce some field extensions over
$\F_q[c]$. Observe that an Artin-Schreier polynomial is either
irreducible or totally split, so at each step of the $p$-descent we
either stay in the same field or we take a degree $p$ extension. This
shows that in the worst case, we have to take an extension of degree
$p^{k-1}$ over $F_q[c]$. The following proposition, which is a
generalisation of \cite[Prop. 26]{Ler97}, states precisely how likely
this case is.

\begin{proposition}
  \label{th:tower}
  Let $E$ be an elliptic curve over $\F_q$, we note $\U_i$ the
  smallest field extension of $\F_q$ such that $E[p^i]\subset
  E(\U_i)$. For any $i\ge1$, either $[\U_{i+1}:\U_i] = p$ or
  $\U_{i+1}=\U_i=\cdots=\U_1$.
\end{proposition}
\begin{proof}
  Observe that the action of the Frobenius $\frobisog$ on $E[p]$ is
  just multiplication by the trace $t$, in fact the equation
  \[\frobisog^2 - [t \bmod p]\circ\frobisog + [q \bmod p] = 0\]
  has two solutions, namely $[t \bmod p]$ and $[0 \bmod p]$, but the
  second can be discarded since it would imply that $\frobisog$ has
  non-trivial kernel.  By lifting this solution, one sees that the
  action of $\frobisog$ on the Tate module $\mathcal{T}_p(E)$ is equal
  to multiplication by some $\tau\in\Z_p$.

  Note $G$ the absolute Galois group of $\F_q$, there is a well known
  action of $G$ on $\mathcal{T}_p(E)$. Since $G$ is generated by the
  Frobenius automorphism of $\F_q$, the restriction of this action to
  $E[p^k]$ is equal to the action (via multiplication) of the subgroup
  of $(\Z/p^k\Z)^\ast$ generated by $\tau_k = \tau \bmod p^k$. Hence
  $[\U_k:\F_q] = \ord(\tau_k)$.

  Then, for any $k>1$, \cite[Corollary 4]{Ler97} applied to
  $\tau_{k+1}=\tau\bmod p^{k+1}$ shows that
  $\ord(\tau_{k+1})=\ord(\tau_k)$ implies
  $\ord(\tau_k)=\ord(\tau_{k-1})$ and this concludes the proof.
\end{proof}

Thus for any elliptic curve there is an $i_0$ such that $[\U_i:\U_1] =
p^{i-i_0}$ for any $i \ge i_0$. This shows that the worst and the
average case coincide since for any fixed curve $[\U_k:\U_1] \in
\Theta(p^k)$ asymptotically. In this situation, one needs
$\Theta(p^kd)$ elements of $\F_p$ to store an element of $\U_k$.

Now the last iteration of the $p$-descent needs to solve an
Artin-Schreier equation in $\U_k$. To do this C2 precomputes the
matrix of the $\F_q$-linear application $(x^q-x):\U_k\rightarrow\U_k$
and its inverse, plus the matrix of the $\F_p$-linear application
$(x^p-x):\F_q\rightarrow\F_q$ and its inverse. The former is the most
expensive one and takes $\Theta(p^{\omega k})$ operations in $\F_q$,
that is $\Omega(p^{\omega k}d) = \Omega(\ell^\omega d)$ operations in
$\F_p$, plus a storage of $\Theta(\ell^2d)$ elements of
$\F_p$. Observe that this precomputation may be used to compute any
other isogeny with domain $E$.

After the precomputation has been done, C2 successively applies the
two inverse matrices; details can be found in
\cite[$\S$2.4]{Cou96}. This costs at least $\Omega(\ell^2d)$.


\paragraph{Interpolation}
The most expensive part of Cauchy interpolation is the polynomial
interpolation phase. In fact, simply representing a polynomial of
degree $p^k-1$ in $\U_k[X]$ takes $\Theta(p^{2k}d)$ elements, thus at
least $\Omega(\ell^2d)$ operations are needed to interpolate unless
special care is taken. This contribution due to arithmetics in $\U_k$
had been underestimated in the complexity analysis of \cite{Cou96},
which gave an estimate of $\Omega(\ell d\log\ell)$ operations for this
phase. We will give more details on interpolation in Section
\ref{sec:C2-AS-FI}.


\paragraph{Recognising the isogeny}
The cost of testing for squareness of the denominator and other tests
is negligible compared to the rest of the algorithm. Nevertheless it
is important to realize that on average half of the $\euler(p^k)$
mappings from $E[p^k]$ to $E'[p^k]$ must be tried before finding the
isogeny, for only one of these mappings corresponds to it. This
implies that the Cauchy interpolation step must be repeated an average
of $\Theta(p^k)$ times, thus contributing a $\Omega(\ell^3d)$ to the
total complexity.

Summing up all the contributions one ends up with the following lower
bound
\begin{equation}
  \label{eq:C2:complexity}
  \Omega(\ell^3d + p^3d)
\end{equation}
plus a precomputation step whose cost is negligible compared to this
one and a space requirement of $\Theta(\ell^2d)$ elements. In the next
sections we will see how to make all these costs drop.


% \subsection{The case $(p,\ell)\ne1$}
% \label{sec:C2:non-prime}
% If we are interested in finding a separable isogeny whose degree is
% not prime to $p$, the best way is to compute the curve $\widetilde{E}$
% such that $E = \widetilde{E}^{(p)}$, then compute an isogeny of degree
% $\ell/p$ between $\widetilde{E}$ and $E'$ and finally compose it with
% the separable $p$-isogeny $V$ from $E$ to $\widetilde{E}$.

% Observe however that C2 can be easily adapted to directly compute such
% an isogeny. In fact let $v=v_p(\ell)$, then $\I(E[p^k]) =
% E'[p^{k-v}]$. All one needs to do in this case is to modify the Cauchy
% interpolation so that it interpolates the rational function that sends
% a generator of $E[p^k]$ over a generator of $E'[p^{k-v}]$ and the
% other points accordingly. The maximum number of trials to do before
% finding the isogeny is $\euler(p^{k-v})$, thus the overall complexity
% is
% \begin{equation}
%   \label{eq:C2:complexity-non-prime}
%   \Omega\left(\frac{\ell^3}{p^v}d + p^3d\right)
%   \;\text{.}
% \end{equation}

% Although this method is less efficient than the first one, it will
% come handy in Section \ref{sec:bounded}.



% Local Variables:
% mode:flyspell
% ispell-local-dictionary:"british"
% mode:TeX-PDF
% TeX-master: "ec-isogeny"
% End:
%
% LocalWords:  Schreier Artin pseudotrace frobenius bivariate Joux Sirvent FFT
% LocalWords:  Couveignes isogenies Schoof isogeny cryptosystems Lercier
% LocalWords:  precomputation arithmetics polylogarithmic Karatsuba precomputes
% LocalWords:  endomorphisms asymptotical
