In this chapter we describe the implementations we made of the
algorithms of the previous chapter and some experimental results.

\section{Implementation of Couveignes' algorithm}
\label{sec:implementation}

We implemented \ctwoasfimc{} as \texttt{C++} programs using the
libraries \texttt{NTL}~\cite{shoup2003ntl} for finite field
arithmetics, \texttt{gf2x}~\cite{gf2x} for fast arithmetics in
characteristic $2$ and \texttt{FAAST} (see
Section~\ref{sec:artin-benchmarks}) for fast arithmetics in
Artin-Schreier towers.  We also have a Magma~\cite{MAGMA} prototype of
the same algorithm, not making use of the fast algorithms of
Chapter~\ref{cha:artin-schr-towers}.  This section mainly deals with
some tricks we implemented in order to speed up the computation.

\subsection{Building \texorpdfstring{$E[p^k]$}{E[pk]} and \texorpdfstring{$E'[p^k]$}{E[pk]}}
\label{sec:impl:torsion}

\paragraph{$p$-Torsion}
For $p\ne2$, \ctwo{} and its variants require to build the extension
$\F_q[c]$ where $c$ is a $(p-1)$-th root of $H_E$. In order to deal with
the lowest possible extension degree, it is a good idea to modify the
curve so that $[\F_q[c]:\F_q]$ is the smallest possible.

$[\F_q[c]:\F_q]$ is invariant under isomorphism, but taking a twist
can save us a quadratic extension. Let $u=c^{-2}$, the curve
\begin{equation*}
  \bar{E} : y^2 = x^3 + a_2ux^2 + a_4u^2x + a_6u^3
\end{equation*}
is defined over $\F_q[c^2]$ and is isomorphic to $E$ over $\F_q[c]$
via $(x,y)\mapsto(\sqrt{u}^2x,\sqrt{u}^3y)$. Its Hasse invariant is
$H_{\bar{E}} = (u)^{\frac{p-1}{2}}H_E = 1$, thus its $p$-torsion
points are defined over $\F_q[c^2]$.

In order to compute the $p^k$-torsion points of $E$ we build
$\F_q[c^2]$, we compute $\bar{P}$ a $p^k$-torsion points of $\bar{E}$
using $p$-descent, then we invert the isomorphism to compute the
abscissa of $P\in E[p^k]$. Since the Cauchy interpolation only needs
the abscissas of $E[p^k]$, this is enough to complete the
algorithm. Scalar multiples of $P$ can be computed without knowledge
of $y(P)$ using \hyperref[rk:montgomery]{Montgomery formulas}.

\pdfmcone{"Kummer surface" -> "Kummer variety".}  Remark that
for $p=2$ we use the same construction in an implicit way since we do
a $p$-descent on the Kummer variety.


\paragraph{$p^k$-Torsion points}
For $p\ne2$ we use Voloch's $p$-descent to compute the $p^k$-torsion
points iteratively as described in Section \ref{sec:C2}. To factor the
Artin-Schreier polynomial \eqref{th:voloch:cover}, we use the
algorithms from Section~\ref{sec:couveignes-algorithm} implemented in
\texttt{FAAST}.

To solve system \eqref{th:voloch:isom} we first compute
\begin{equation*}
  V(x,y) = \left(\frac{g(x)}{h^2(x)}, 
    sy\left(\frac{g(x)}{h^2(x)}\right)'\right)
\end{equation*}
through \titleref{sec:velu-formulas}.\footnote{Vélu formulas compute
  this isogeny up to an indeterminacy on the sign of the ordinate, the
  actual value of $s$ must be determined by composing $V$ with
  $\frobisog$ and verifying that it corresponds to $[p]$ by trying
  some random points.} Recall that we work on a curve having Hasse
invariant $1$, system \eqref{th:voloch:isom} can then be rewritten
\begin{equation*}
  \left\{
    \begin{aligned}
      \tilde{x}(P) &= \frac{g(x)}{h^2(x)}\\
      \tilde{y}(P) &= sy\left(\frac{g(x)}{h^2(x)}\right)'\\
      \tilde{z}(P) &= -2y\frac{h'(x)}{h(x)}
    \end{aligned}
  \right.
\end{equation*}
where $P$ is the point on the cover $C$ that we want to pull back
($\tilde{x}(P)$, $\tilde{y}(P)$ and $\tilde{z}(P)$ are just its
coordinates). After some substitutions this is equivalent to
\begin{equation*}
  \left\{
    \begin{aligned}
      \tilde{x}(P)h^2(x) - g(x) &= 0\\
      \left(\tilde{x}(P)h^2(x) - g(x) - \frac{\tilde{y}(P)}{s\tilde{x}(P)}h^2(x)\right)' &= 0
    \end{aligned}
  \right.
\end{equation*}
Then a solution in $x$ to this system is given by the GCD of the two
equations. Remark that proposition \ref{th:voloch} ensures there is
one unique solution. These formulas are slightly more efficient than
the ones in \cite[$\S$6.2]{lercier-algorithmique}.

For $p=2$ we use the library \texttt{FAAST} (for solving
Artin-Schreier equations) on top of \texttt{gf2x} (for better
performance). There is nothing special to remark about the
$2$-descent.


\subsection{Cauchy interpolation and loop}
\label{sec:impl:cauchy}
The polynomial interpolation step is done as described in Section
\ref{sec:C2-AS-FI}. As a result of this implementation, the polynomial
interpolation algorithm was added to the library \texttt{FAAST}.

The rational fraction reconstruction is implemented using a fast XGCD
algorithm on top of \texttt{NTL} and \texttt{gf2x}. This algorithm was
added to \texttt{FAAST} too.

The loop uses modular composition as in Section~\ref{sec:C2-AS-FI-MC}
in order to minimise the number of interpolations. The timings in the
next section clearly show that this non-asymptotically-optimal variant
performs much faster in practice.

To check that the rational fractions are isogenies we test their
degrees, that their denominator is a square and that they act as group
morphisms on a fixed number of random points. All these checks take a
negligible amount of time compared to the rest of the algorithm.


\subsection{Parallelisation of the loop}
\label{parallel}

The most expensive step of \ctwoasfimc{}, in theory as well as in
practice, is the final loop over the points of $E'[p^k]$. Fortunately,
this phase is very easy to parallelise with very little overhead.

Let $n$ be the number of processors we wish to parallelise on, suppose
that $[\U_k:\F_q]$ is maximal, then we make only one interpolation
followed by $\euler(p^k)/2$ modular compositions.\footnote{If
  $[\U_k:\F_q]$ is not maximal, the parallelisation is
  straightforward: we simply send one interpolation to each processor
  in turn.} We set $m=\left\lfloor\frac{\euler(p^k)}{2n}\right\rfloor$
and we compute the action of $\frobisog^{m}$ on $E[p^k]$ as in
Section~\ref{sec:C2-AS-FI-MC}:
\begin{equation*}
  F^{(m)}(X) = F(X) \circ \cdots \circ F(X) \bmod T(X)
  \;\text{,}
\end{equation*}
this can be done with $\Theta(\log m)$ modular compositions via a
binary square-and-multiply approach as in
Section~\ref{sec:modular-composition}.

Then we compute the $n$ polynomials
\begin{equation*}
  A_{mi}(X) = A_{m(i-1)}(X) \circ F^{(m)}(X) \bmod T(X)
\end{equation*}
and distribute them to the $n$ processors so that they each work on a
separate slice of the $A_i$'s. The only overhead is $\Theta(\log
(\ell/n))$ modular compositions with coefficients in $\F_q$, this is
acceptable in most cases.

\section{Implementation of \ctwoud{}}
\label{sec:implementation-c2-ud}
We modified our \texttt{C++} implementation of \ctwoasfimc{} to obtain
two variants of \ctwoud{}.

The first one takes an integer $k$ and looks for all isogenies of
degree $p^c\ell$ with $\ell$ prime to $p$, $\ell<\euler(p^k)/4$
and $c$ arbitrary. This is done by slightly modifying the modular
composition step of Section~\ref{sec:C2-AS-FI-MC}. Suppose we know an
interpolating polynomial $A_0$, that we view as a morphisms $E[p^k]\ra
E'[p^k]$ such that
\begin{equation}
  \label{eq:184}
  A_0\circ[n](P) = [n](P')
  \quad\text{for any $n$.}
\end{equation}
Then we compose with the Frobenius isogeny $\frobisog$
\begin{equation}
  \label{eq:185}
  A_1 \eqdef \frobisog\circ A_0:E[p^k]\ra {E'}^{(p)}[p^k]
  \text{,}
\end{equation}
where ${E'}^{(p)}$ is the curve
\begin{equation}
  \label{eq:186}
  E^{(p)}\;:\; y^2 = x^3+{a'}^px + {b}'^p
  \text{.}
\end{equation}
So $A_1$ is one of the polynomials that Couveignes algorithm computes
when looking for an isogeny between $E$ and ${E'}^{(p)}$. If $q=p^d$,
iterating $d$ times this construction, we fall back on $E'$, as we
would have if we had directly applied the
\hyperref[sec:curves-over-finite]{Frobenius automorphism} as in
Section~\ref{sec:C2-AS-FI-MC}. Thus, paying an additional factor of
$\log_p q$, we can compute any isogeny of degree $p^c\ell$ with
arbitrary $c$ and $\ell$ bounded as before.

When $\log_pq$ is large, the previous variant becomes unpractical. We
implemented a second variant using the algorithm described in
Section~\ref{sec:C2:non-prime}, this allows to compute any isogeny of
degree $\ell\le\euler(p^k)/4$, even if $p$ divides $\ell$. The
asymptotic cost of this variant is the same as one run of \ctwoasfimc{},
because the search for $\ell$ prime to $p$ dominates.



\section{Implementation of Lercier-Sirvent}
We implemented a Magma prototype of \titleref{alg:bmss}
and \titleref{alg:le-si}. In both cases we did
not take Remark~\ref{rk:bmss} into account, and only implemented the
variant using rational fraction reconstruction. We used Magma native
support for $p$-adics to construct the field $\Q_q$.

Both implementations differ from the description we gave in
Sections~\ref{sec:bmss} and~\ref{sec:lercier-sirvent} in that they
only take the degree $\ell$ and the equation of $E$ as input, and
compute an $\F_q$-isogenous curve, if it exists, by factoring the
modular polynomial in $\F_q[X]$.

\pdfmctwo{Atkin's MODULAR polynomial. Not "canonical".}
Instead of the classical modular polynomials $\Modpol_\ell$ we used
Atkin's modular polynomials $\Modpol^\ast_\ell$ since they have
smaller coefficients and degree; this does not change the other steps
of the algorithm.

The modular polynomials were not computed on the fly as suggested in
Section~\ref{sec:lercier-sirvent}, instead they were taken from the
tables precomputed in Magma. This implies that the implementation has
an asymptotic complexity cubic in $\ell$, however we will see that
even this implementation behaves well in practice.


% Local Variables:
% mode:flyspell
% ispell-local-dictionary:"american"
% mode:TeX-PDF
% TeX-master: "../these"
% mode:reftex
% End:
%
% LocalWords:  Schreier Artin pseudotrace Frobenius bivariate Joux Sirvent FFT
% LocalWords:  Couveignes isogenies Schoof isogeny cryptosystems Lercier Hasse
% LocalWords:  precomputation arithmetics polylogarithmic Karatsuba precomputes
% LocalWords:  endomorphisms 
